---
title: "Image Feature Extraction Assignment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Setup — telling R where your files live

Before anything else, just set these two paths to match wherever you saved the images folder and the CSV on your computer. Everything else in the script will use these.

```{r paths}
# ---- CHANGE THESE TO MATCH YOUR SETUP ----
images_dir <- "images"       # folder containing all the .png files
labels_csv  <- "labels.csv"  # the csv with brand / product-type columns
# ------------------------------------------
```

---

# Question 1 — Loading the libraries

We load four specialist image-processing packages plus two modelling packages. `pixmap` handles raw pixel data, `imager` does edge detection and general image manipulation, `OpenImageR` provides Gabor texture filters, and `colordistance` builds colour histograms. `e1071` gives us the SVM classifier and `MLmetrics` gives accuracy and F1 scoring.

```{r libraries}
library(pixmap)         # colour / pixel-level image representation
library(imager)         # edge detection (Canny) and general image manipulation
library(OpenImageR)     # Gabor texture filters
library(colordistance)  # colour histogram extraction (RGB & HSV)
library(e1071)          # SVM
library(MLmetrics)      # accuracy / F1 / recall
```

---

# Question 2 — Loading all 2 184 images

We collect every PNG path from the images folder, load each one with `imager`, and strip the alpha (transparency) channel if present — keeping only the three RGB channels. Feature-extraction functions downstream expect exactly three channels and will fail or misbehave if a fourth is present.

```{r load-images}
image_paths <- list.files(images_dir, pattern = "\\.png$",
                           full.names = TRUE, recursive = FALSE)

cat("Total images found:", length(image_paths), "\n")

images <- lapply(image_paths, function(p) {
  img <- load.image(p)
  if (dim(img)[4] == 4) {
    img <- img[, , , 1:3, drop = FALSE]
    img <- as.cimg(img)
  }
  img
})

cat("Channels in image 1:", dim(images[[1]])[4], "\n")
cat("Dimensions (W x H x D x C):", paste(dim(images[[1]]), collapse = " x "), "\n")
```

```{r show-example-image, fig.cap="Example clothing image after alpha removal"}
plot(images[[1]], main = "Example image — 3 channels only")
```

---

# Question 3 — Loading the labels

We read the CSV that records what brand and product type each image belongs to. Every row corresponds to one image in the same order as `image_paths`. As features are extracted in later questions they are appended as new columns, so by Question 7 this data frame holds both the features and the targets.

```{r load-labels}
labels <- read.csv(labels_csv, stringsAsFactors = FALSE)

cat("Rows:", nrow(labels), "\n")
cat("Columns:", paste(names(labels), collapse = ", "), "\n")
head(labels)

cat("\nBrand counts:\n");        print(table(labels$brand))
cat("\nProduct type counts:\n"); print(table(labels$productType))
```

---

# Question 4 — Colour histograms (RGB and HSV)

We use `getImageHist()` to build histograms with 3 bins per channel, giving 3³ = 27 values per colour space per image. The helper returns the `Pct` column as a plain numeric vector. Both the RGB and HSV matrices are appended to `labels`.

```{r colour-histograms}
n_bins <- 3   # bins per channel → 3^3 = 27 values per colour space

get_hist_vec <- function(path, color_space) {
  h <- getImageHist(path,
                    bins        = n_bins,
                    color.space = color_space,
                    plotting    = FALSE)
  as.numeric(h$Pct)
}

# RGB histograms
cat("Extracting RGB histograms...\n")
rgb_list <- lapply(seq_along(image_paths), function(i) {
  if (i %% 200 == 0) cat("  processed", i, "/", length(image_paths), "\n")
  get_hist_vec(image_paths[i], "rgb")
})
rgb_matrix <- do.call(rbind, rgb_list)
colnames(rgb_matrix) <- paste0("rgb_", seq_len(ncol(rgb_matrix)))

# HSV histograms
cat("Extracting HSV histograms...\n")
hsv_list <- lapply(seq_along(image_paths), function(i) {
  if (i %% 200 == 0) cat("  processed", i, "/", length(image_paths), "\n")
  get_hist_vec(image_paths[i], "hsv")
})
hsv_matrix <- do.call(rbind, hsv_list)
colnames(hsv_matrix) <- paste0("hsv_", seq_len(ncol(hsv_matrix)))

labels <- cbind(labels, rgb_matrix, hsv_matrix)

cat("Feature columns so far:", ncol(labels), "\n")
```

A colour histogram counts how many pixels fall into each colour bucket. It does not care where colours appear in the image — just how much of each is present overall. RGB captures raw colour mixture; HSV separates hue (the actual colour tone) from saturation and brightness, giving the classifier a complementary view of the same palette.

---

# Question 5 — Edge detection and line detection

We convert each image to grayscale (Canny requires a single-channel input), run `cannyEdges()` to get a binary edge map, then apply `hough_line()` to detect straight lines. Only lines scoring at or above the 99.5th percentile are kept; their count per image is the feature added to `labels`.

```{r edge-line-detection}
n_lines_vec <- numeric(length(images))

cat("Running edge + line detection...\n")
for (i in seq_along(images)) {
  if (i %% 200 == 0) cat("  processed", i, "/", length(images), "\n")

  gray_img  <- grayscale(images[[i]])
  edges     <- cannyEdges(gray_img)
  hl        <- hough_line(edges, ntheta = 800)
  scores    <- as.numeric(hl)
  threshold <- quantile(scores, probs = 0.995)
  n_lines_vec[i] <- sum(scores >= threshold)
}

labels$n_important_lines <- n_lines_vec

cat("Line feature added. Sample values:", head(n_lines_vec, 10), "\n")
```

**Step by step through the pipeline:**

- **Grayscale conversion** — `grayscale()` collapses the three colour channels into one intensity channel. Both Canny and Hough work on brightness only, so colour information is not needed and would just complicate the computation.
- **Canny edge detection** — `cannyEdges()` produces a binary map where each pixel is either an edge (1) or not (0). It does this by finding where pixel brightness changes sharply, then thinning those regions down to single-pixel-wide lines.
- **Hough line detection** — `hough_line(edges, ntheta = 800)` scans the edge map looking for straight lines. The `ntheta = 800` argument means it tests 800 different angles evenly spaced between 0° and 180°, giving finer angular resolution and more precise line orientations.
- **Percentile threshold** — `hough_line()` returns a matrix where each cell value is the score for a (distance, angle) pair. We flatten it with `as.numeric()` and use `quantile(..., probs = 0.995)` to find the top 0.5 % threshold. Counting cells at or above that threshold keeps only the strongest, most consistently supported lines and discards weak or noisy detections.
- **Single numeric feature per image** — the count of lines above that threshold is stored as one number in `n_lines_vec[i]`. After the loop, `labels$n_important_lines <- n_lines_vec` appends it as a new column so every row in the label dataset now has a value representing how many strong straight lines were found in that image. A garment with sharp geometric structure (e.g. a striped shirt or tailored jacket) will score higher than a plain or loosely shaped item.
- **Why edge detection is a necessary preprocessing step for line detection** — the Hough transform works by having each edge pixel vote for every line it could lie on. It needs a binary input where pixels are either definitively an edge or not. If you skipped Canny and fed in raw grayscale, every pixel would carry a continuous brightness value and there would be nothing to vote with — just a smooth gradient. Canny reduces the image to a sparse map of sharp boundaries, which gives Hough a clean, unambiguous set of points to work from and makes the detected lines meaningful rather than artefacts of overall brightness.

---

# Question 6 — Texture features with Gabor filters

`gabor_feature_extraction()` runs a bank of 2-D Gabor filters at multiple scales and orientations. For each scale we summarise the filter response with its mean (how much of that texture is present) and standard deviation (how evenly spread it is). The resulting matrix is appended to `labels`.

```{r gabor-texture}
# probe one image to discover the number of scales
test_gabor <- gabor_feature_extraction(
  image = images[[1]], scales = NULL, orientations = NULL, gabor_type = "2d"
)
n_scales <- length(test_gabor$gabor_features_real)
cat("Number of Gabor scales:", n_scales, "\n")

gabor_col_names <- c(paste0("gabor_mean_scale", seq_len(n_scales)),
                     paste0("gabor_sd_scale",   seq_len(n_scales)))

cat("Extracting Gabor features...\n")
gabor_list <- lapply(seq_along(images), function(i) {
  if (i %% 200 == 0) cat("  processed", i, "/", length(images), "\n")
  g <- gabor_feature_extraction(
    image = images[[i]], scales = NULL, orientations = NULL, gabor_type = "2d"
  )
  c(sapply(g$gabor_features_real, mean),
    sapply(g$gabor_features_real, sd))
})

gabor_matrix <- do.call(rbind, gabor_list)
colnames(gabor_matrix) <- gabor_col_names

labels <- cbind(labels, gabor_matrix)

cat("Total feature columns now:", ncol(labels), "\n")
```

A Gabor filter is a wave pattern slid across the image; it responds strongly wherever pixel intensities repeat at the filter's frequency and orientation — stitching, weave patterns, smooth fabric. The mean captures how much of that texture is present globally; the SD captures how patchy or uniform it is across the garment.

---

# Question 7 — Classification models

We do an 80/20 train-test split, scale the features to zero mean and unit variance (SVMs are sensitive to scale), then train one SVM predicting brand and another predicting product type. Both are evaluated with accuracy, macro F1, and a confusion matrix.

```{r modelling}
set.seed(42)

label_cols   <- c("brand", "productType")
feature_cols <- setdiff(names(labels), label_cols)

complete_rows  <- complete.cases(labels[, feature_cols])
cat("Rows with complete features:", sum(complete_rows), "/", nrow(labels), "\n")
modelling_data <- labels[complete_rows, ]

n         <- nrow(modelling_data)
train_idx  <- sample(seq_len(n), size = floor(0.8 * n))
train_data <- modelling_data[ train_idx, ]
test_data  <- modelling_data[-train_idx, ]

cat("Training rows:", nrow(train_data), " | Test rows:", nrow(test_data), "\n")

X_train    <- as.matrix(train_data[, feature_cols])
X_test     <- as.matrix(test_data[,  feature_cols])
X_train_sc <- scale(X_train)
X_test_sc  <- scale(X_test,
                    center = attr(X_train_sc, "scaled:center"),
                    scale  = attr(X_train_sc, "scaled:scale"))
```

```{r model-brand}
y_brand_train <- as.factor(train_data$brand)
y_brand_test  <- as.factor(test_data$brand)

brand_model <- svm(X_train_sc, y_brand_train, kernel = "linear", cost = 1)
brand_preds <- predict(brand_model, X_test_sc)

cat("\n=== BRAND MODEL ===\n")
cat("Accuracy:", Accuracy(brand_preds, y_brand_test), "\n")
cat("Macro F1:", mean(sapply(levels(y_brand_test), function(cls) {
  F1_Score(as.integer(y_brand_test == cls), as.integer(brand_preds == cls))
}), na.rm = TRUE), "\n")
```

```{r model-product-type}
y_type_train <- as.factor(train_data$productType)
y_type_test  <- as.factor(test_data$productType)

type_model <- svm(X_train_sc, y_type_train, kernel = "linear", cost = 1)
type_preds <- predict(type_model, X_test_sc)

cat("\n=== PRODUCT TYPE MODEL ===\n")
cat("Accuracy:", Accuracy(type_preds, y_type_test), "\n")
cat("Macro F1:", mean(sapply(levels(y_type_test), function(cls) {
  F1_Score(as.integer(y_type_test == cls), as.integer(type_preds == cls))
}), na.rm = TRUE), "\n")
```

```{r confusion-matrices}
cat("\nBrand confusion matrix:\n")
print(table(Predicted = brand_preds, Actual = y_brand_test))

cat("\nProduct type confusion matrix:\n")
print(table(Predicted = type_preds, Actual = y_type_test))
```

Product type is likely the easier task — a jacket, a pair of jeans, and a dress differ dramatically in shape, colour, and texture, and those differences are consistent across brands. Brand is harder because different brands sell the same product types and often use similar palettes; the model must pick up subtler cues like cut or preferred fabric texture. Expect accuracy and F1 to be noticeably lower for brand.
