---
title: "Image Feature Extraction Assignment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Setup — telling R where your files live

Before anything else, just set these two paths to match wherever you saved the images folder and the CSV on your computer. Everything else in the script will use these.

```{r paths}
# ---- CHANGE THESE TO MATCH YOUR SETUP ----
images_dir <- "images"   # folder containing all the .png files
labels_csv  <- "labels.csv"  # the csv with brand / product-type columns
# ------------------------------------------
```

---

# Question 1 — Loading the libraries

Here we load the four image-processing libraries the assignment needs. Each one brings different tools to the table, so it makes sense to think of them as specialists.

```{r libraries}
library(pixmap)         # colour / pixel-level image representation
library(imager)         # edge detection (Canny) and general image manipulation
library(OpenImageR)     # Gabor texture filters
library(colordistance)  # colour histogram extraction (RGB & HSV)

# extra libraries needed for modelling and evaluation
library(e1071)          # SVM
library(MLmetrics)      # accuracy / F1 / recall
```

So to break it down quickly:

- **pixmap** — mainly for *colour* (reading and working with raw pixel data)
- **imager** — mainly for *shape* (edge detection, contour stuff)
- **OpenImageR** — mainly for *texture* (Gabor filters)
- **colordistance** — mainly for *colour* (building colour histograms)

---

# Question 2 — Loading all 2 184 images

We grab every PNG file from the images folder, load each one with `imager`, and immediately drop the alpha (transparency) channel if it's there, keeping only the three RGB channels. At the end we do a quick sanity check and show one image.

```{r load-images}
# collect all png paths
image_paths <- list.files(images_dir, pattern = "\\.png$",
                          full.names = TRUE, recursive = FALSE)

cat("Total images found:", length(image_paths), "\n")

# load and strip alpha channel
images <- lapply(image_paths, function(p) {
  img <- load.image(p)
  # imager stores channels in the 4th dimension; keep only channels 1-3
  if (dim(img)[4] == 4) {
    img <- img[, , , 1:3, drop = FALSE]
    img <- as.cimg(img)
  }
  img
})

# verify dimensions on the first image
cat("Channels in image 1:", dim(images[[1]])[4], "\n")
cat("Dimensions (W x H x D x C):", paste(dim(images[[1]]), collapse = " x "), "\n")
```

```{r show-example-image, fig.cap="Example clothing image after alpha removal"}
plot(images[[1]], main = "Example image — 3 channels only")
```

Removing the alpha channel matters because most of the feature-extraction functions we use later expect exactly three channels (R, G, B). If there's a fourth transparency channel hanging around, those functions either crash or produce garbage, so we get rid of it up front.

---

# Question 3 — Loading the labels

We read the CSV, take a peek at what columns are there, and confirm we have a brand column and a product-type column to predict.

```{r load-labels}
labels <- read.csv(labels_csv, stringsAsFactors = FALSE)

# quick look at the data
cat("Rows:", nrow(labels), "\n")
cat("Columns:", paste(names(labels), collapse = ", "), "\n")
head(labels)

# check class balance (adjust column names if yours differ)
cat("\nBrand counts:\n");   print(table(labels$brand))
cat("\nProduct type counts:\n"); print(table(labels$product_type))
```

The idea is simple: every row in `labels` corresponds to one image (in the same order as the files in the folder). As we extract features from each image we'll create new columns — one per feature — and tack them onto this data frame. By the end we have one big table where each row is an image with its features *and* its labels, ready to feed into a model.

---

# Question 4 — Extracting Colour Histograms (RGB and HSV)

## What this question asks

This question requires us to:

1. Import the **label dataset** (brand and product type) using `read.csv()`.
2. Extract **RGB colour histograms** from all images using `getImageHist()` from the `colordistance` package.
3. Repeat the extraction in **HSV colour space**.
4. Use exactly **3 bins per channel** in both cases, producing 3³ = **27 features** per image per colour space.
5. Ensure all features are stored as plain **numeric values**.
6. Build one **RGB feature matrix** and one **HSV feature matrix**, then **append both to the label dataset**.
7. Explain what colour histograms capture about an image.

---

## Step 4a — Import the label dataset

We first load the CSV that contains one row per image with the ground-truth labels we want to predict: **brand** and **product type**. These will be the target columns once we attach the image features.

```{r load-labels-q4}
labels <- read.csv(labels_csv, stringsAsFactors = FALSE)

cat("Rows:", nrow(labels), "\n")
cat("Columns:", paste(names(labels), collapse = ", "), "\n")
head(labels)

cat("\nBrand counts:\n");        print(table(labels$brand))
cat("\nProduct type counts:\n"); print(table(labels$productType))
```

Each row in `labels` corresponds to one image file (in the same order the files appear on disk). As we extract features from each image we will add new columns to this data frame — one per feature — so that by the end every row holds both the numeric image description **and** the label to predict. This combined table is what gets fed into the classifier.

---

## Step 4b — Extract RGB colour histograms

`getImageHist()` reads an image directly from its file path, divides each colour channel into the requested number of equal-width bins, and counts what fraction of pixels fall in each bin. Setting `bins = 3` gives 3 bins per channel, so 3 × 3 × 3 = **27 bins** in total.

The function returns a data frame; we pull out the `Pct` column (the per-bin pixel fractions) and coerce it to a plain numeric vector so every row of the resulting matrix is in a consistent numeric format.

```{r rgb-histograms}
n_bins <- 3   # 3 bins per channel → 27 features per image

# helper: run getImageHist and return a plain numeric vector
get_hist_vec <- function(path, color_space) {
  h <- getImageHist(path,
                    bins        = n_bins,
                    color.space = color_space,
                    plotting    = FALSE)
  as.numeric(h$Pct)   # pixel fraction for each of the 27 bins
}

# extract for every image
cat("Extracting RGB histograms...\n")
rgb_list <- lapply(seq_along(image_paths), function(i) {
  if (i %% 200 == 0) cat("  processed", i, "/", length(image_paths), "\n")
  get_hist_vec(image_paths[i], "rgb")
})

# stack into one matrix: rows = images, columns = histogram bins
rgb_matrix <- do.call(rbind, rgb_list)
colnames(rgb_matrix) <- paste0("rgb_", seq_len(ncol(rgb_matrix)))

cat("RGB matrix dimensions:", nrow(rgb_matrix), "x", ncol(rgb_matrix), "\n")
cat("All numeric?", all(apply(rgb_matrix, 2, is.numeric)), "\n")
```

---

## Step 4c — Extract HSV colour histograms

We repeat exactly the same procedure but pass `color.space = "hsv"`. The HSV space decomposes colour into **Hue** (the actual colour tone), **Saturation** (how vivid it is), and **Value** (how bright or dark). This gives a different angle on colour than raw RGB, which is why both are useful.

```{r hsv-histograms}
cat("Extracting HSV histograms...\n")
hsv_list <- lapply(seq_along(image_paths), function(i) {
  if (i %% 200 == 0) cat("  processed", i, "/", length(image_paths), "\n")
  get_hist_vec(image_paths[i], "hsv")
})

hsv_matrix <- do.call(rbind, hsv_list)
colnames(hsv_matrix) <- paste0("hsv_", seq_len(ncol(hsv_matrix)))

cat("HSV matrix dimensions:", nrow(hsv_matrix), "x", ncol(hsv_matrix), "\n")
cat("All numeric?", all(apply(hsv_matrix, 2, is.numeric)), "\n")
```

---

## Step 4d — Append both feature matrices to the label dataset

We use `cbind()` to attach the 27 RGB columns and the 27 HSV columns to the right of the labels data frame. Because every matrix has exactly as many rows as `labels` has rows (one per image, in the same order), the join is purely positional — no key column needed.

```{r append-colour-features}
labels <- cbind(labels, rgb_matrix, hsv_matrix)

cat("Total columns in labels after appending colour features:", ncol(labels), "\n")
cat("First few column names:", paste(names(labels)[1:10], collapse = ", "), "...\n")
```

---

## What information do colour histograms capture?

A colour histogram answers one question: **how much of each colour is present in the image?** It divides each colour channel into bins and records what fraction of pixels fall in each bin.

| Colour space | What the three channels represent |
|---|---|
| **RGB** | How much **red**, **green**, and **blue** light makes up the overall palette |
| **HSV** | **Hue** — the colour tone (red vs. blue vs. green); **Saturation** — how vivid vs. washed-out; **Value** — how bright vs. dark |

With 3 bins per channel and 27 bins total, the histogram gives a coarse but compact summary of the colour palette. Critically, it is **position-agnostic**: it does not matter where in the image the red pixels are, only how many there are. This makes histograms robust to small shifts, rotations, and scale changes, and fast to compute across thousands of images.

For clothing classification specifically, colour histograms are informative because different brands and product types tend to have characteristic colour palettes — denim jeans skew blue, formal shirts tend towards whites and light blues, and some brands have distinctive signature colours. These 54 features (27 RGB + 27 HSV) therefore give the classifier a meaningful colour fingerprint for each garment before any shape or texture information is added.

---

# Question 5 — Edge detection and line detection

First we convert each image to grayscale (Canny only works on single-channel images), then detect edges with `cannyEdges()`, then run `hough_line()` on those edges. We only keep lines whose score hits the **99.5th percentile** threshold, and the feature is simply the count of those important lines.

```{r edge-line-detection}
n_lines_vec <- numeric(length(images))

cat("Running edge + line detection...\n")
for (i in seq_along(images)) {
  if (i %% 200 == 0) cat("  processed", i, "/", length(images), "\n")

  # grayscale
  gray_img <- grayscale(images[[i]])

  # Canny edges
  edges <- cannyEdges(gray_img)

  # Hough line transform (theta = 800 as specified)
  hl <- hough_line(edges, theta = 800)

  # keep only lines in the top 0.5% by score
  threshold <- quantile(hl$score, probs = 0.995)
  n_lines_vec[i] <- sum(hl$score >= threshold)
}

labels$n_important_lines <- n_lines_vec

cat("Line feature added. Sample values:", head(n_lines_vec, 10), "\n")
```

Edge detection is a prerequisite for line detection because Hough transform works on a binary edge map — it looks for straight sequences of "edge pixels". If you fed it the raw greyscale image it would just see gradients and brightness variation, not clean lines. Canny sharpens that raw pixel soup into crisp edges that Hough can actually count as lines.

---

# Question 6 — Texture features with Gabor filters

`gabor_feature_extraction()` runs a bank of filters at different scales and orientations. We grab the `gabor_features_real` list (one filtered image per scale) and summarise each with its **mean** and **standard deviation**, giving us a compact set of texture numbers per image.

```{r gabor-texture}
# run one image first to find out how many scales we get
test_gabor <- gabor_feature_extraction(
  image        = images[[1]],
  scales       = NULL,     # default
  orientations = NULL,     # default
  gabor_type   = "2d"
)
n_scales <- length(test_gabor$gabor_features_real)
cat("Number of Gabor scales:", n_scales, "\n")

# feature names: mean and sd for each scale
gabor_col_names <- c(
  paste0("gabor_mean_scale", seq_len(n_scales)),
  paste0("gabor_sd_scale",   seq_len(n_scales))
)

# extract for all images
cat("Extracting Gabor features...\n")
gabor_list <- lapply(seq_along(images), function(i) {
  if (i %% 200 == 0) cat("  processed", i, "/", length(images), "\n")

  g <- gabor_feature_extraction(
    image        = images[[i]],
    scales       = NULL,
    orientations = NULL,
    gabor_type   = "2d"
  )

  means <- sapply(g$gabor_features_real, mean)
  sds   <- sapply(g$gabor_features_real, sd)
  c(means, sds)
})

gabor_matrix <- do.call(rbind, gabor_list)
colnames(gabor_matrix) <- gabor_col_names

labels <- cbind(labels, gabor_matrix)

cat("Total feature columns now:", ncol(labels), "\n")
```

The Gabor filter is essentially a tiny wave pattern you slide across the image. It responds strongly wherever there's a repeating texture at a particular scale and direction — think weave patterns in fabric, stitching, or smooth surfaces. The mean tells you how much of that texture is present overall; the SD tells you how varied or localised it is. Together they give the model a sense of how coarse or fine the surface of the clothing looks.

---

# Question 7 — Classification models

We pull together all the features, do an 80/20 train-test split, and train two SVMs — one predicting **brand**, one predicting **product type**. Then we evaluate with accuracy, recall, and F1.

```{r modelling}
set.seed(42)

# identify feature columns (everything except the label columns)
label_cols   <- c("brand", "product_type")    # adjust if your CSV uses different names
feature_cols <- setdiff(names(labels), label_cols)

# remove any rows that ended up with NA features
complete_rows <- complete.cases(labels[, feature_cols])
cat("Rows with complete features:", sum(complete_rows), "/", nrow(labels), "\n")
modelling_data <- labels[complete_rows, ]

# 80 / 20 split
n          <- nrow(modelling_data)
train_idx  <- sample(seq_len(n), size = floor(0.8 * n))
train_data <- modelling_data[ train_idx, ]
test_data  <- modelling_data[-train_idx, ]

cat("Training rows:", nrow(train_data), " | Test rows:", nrow(test_data), "\n")

# feature matrices
X_train <- as.matrix(train_data[, feature_cols])
X_test  <- as.matrix(test_data[, feature_cols])

# scale features (SVMs are sensitive to scale)
X_train_sc <- scale(X_train)
scaler_mean <- attr(X_train_sc, "scaled:center")
scaler_sd   <- attr(X_train_sc, "scaled:scale")
X_test_sc   <- scale(X_test, center = scaler_mean, scale = scaler_sd)
```

```{r model-brand}
# --- Brand model ---
y_brand_train <- as.factor(train_data$brand)
y_brand_test  <- as.factor(test_data$brand)

brand_model <- svm(X_train_sc, y_brand_train,
                   kernel = "linear", cost = 1,
                   probability = FALSE)

brand_preds <- predict(brand_model, X_test_sc)

cat("\n=== BRAND MODEL ===\n")
cat("Accuracy:", Accuracy(brand_preds, y_brand_test), "\n")
cat("Macro F1:", mean(sapply(levels(y_brand_test), function(cls) {
  F1_Score(y_true = as.integer(y_brand_test == cls),
           y_pred = as.integer(brand_preds   == cls))
}), na.rm = TRUE), "\n")
```

```{r model-product-type}
# --- Product type model ---
y_type_train <- as.factor(train_data$product_type)
y_type_test  <- as.factor(test_data$product_type)

type_model <- svm(X_train_sc, y_type_train,
                  kernel = "linear", cost = 1,
                  probability = FALSE)

type_preds <- predict(type_model, X_test_sc)

cat("\n=== PRODUCT TYPE MODEL ===\n")
cat("Accuracy:", Accuracy(type_preds, y_type_test), "\n")
cat("Macro F1:", mean(sapply(levels(y_type_test), function(cls) {
  F1_Score(y_true = as.integer(y_type_test == cls),
           y_pred = as.integer(type_preds   == cls))
}), na.rm = TRUE), "\n")
```

```{r confusion-matrices}
cat("\nBrand confusion matrix:\n")
print(table(Predicted = brand_preds, Actual = y_brand_test))

cat("\nProduct type confusion matrix:\n")
print(table(Predicted = type_preds, Actual = y_type_test))
```

**Discussion.** Product type (e.g. t-shirt vs trousers vs jacket) is likely the easier prediction task. Shape features like edge counts and line patterns, together with colour histograms, provide pretty strong visual signals for coarse category differences — a pair of jeans genuinely looks very different from a dress. Brand, on the other hand, can share the same product types and colours across brands, so the model has to pick up on more subtle visual cues (logo placement, cut, palette preferences). Expect accuracy to be noticeably lower for brand than for product type.
```
